---
title: "OSplines-Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OSplines-Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", fig.height = 3, fig.width = 5, margins = TRUE
)
knitr::knit_hooks$set(margins = function(before, options, envir) {
  if (!before) {
    return()
  }
  graphics::par(mar = c(1.5 + 0.9, 1.5 + 0.9, 0.2, 0.2), mgp = c(1.45, 0.45, 0), cex = 1.25, bty = "n")
})
```


```{r setup}
library(OSplines)
library(tidyverse)
```

# Idea of O-Spline approximation
Let $f(t)$ be the unknown function that we want to infer, and assume it is already known that $f$ is at least $p$-1 times continuously differentiable.

Consider the smoothing model:
$$\frac{\partial^p{f}(t)}{\partial t^p} = \sigma_sW(t),$$
with the boundary (initial) conditions that $\frac{\partial^q{f}(0)}{\partial t^q} = 0$ for all $0\leq q <p$.
Here $W(t)$ is the standard Gaussian white noise process, or can be viewed as the distributional derivative of the standard Brownian motion.

This kind of smoothing model is referred to as Integrated Wiener's process (IWP), and both the O-spline method and the commonly used continuous RW2 method consider its finite dimensional approximation in the following form:
$$\tilde{f}(t) = \sum_{i=1}^{k} w_i\varphi_i(t),$$
where $\boldsymbol{w} = (w_1, ..., w_k)^T$ is the vector of Gaussian coefficients, and $\varphi_i$ is basis function defined at the $i$th knot.

The main target of inference is therefore the Gaussian coefficients vector $\boldsymbol{w}$, which will have a diagonal precision matrix under the O-spline construction.

# Example of Munich Data

## Data and Model

We will illustrate the use of O-spline using the Munich rent data from INLA's Rpackage:

```{r}
### A model with two IWP and two Fixed effects:
data <- INLA::Munich %>% select(rent, floor.size, year, location)
data$score <- rnorm(n = nrow(data))
data$year <- seq(1900, 2010, length.out = nrow(data))
func <- function(x) {sin(0.5*x)}
inv_logit_func <- function(x) {exp(x)/(1 + exp(x))}
data$response = rbinom(n = nrow(data), size = 1, prob = inv_logit_func(func(data$year)))
head(data, n = 5)
polyOrder1 <- 2 ## Assume f(floor.size) is second order IWP
polyOrder2 <- 3 ## Assume f(year) is third order IWP
```

Given the set of knots, we can easily construct the basis function $\varphi_i$ and the precision matrix of $\boldsymbol{w}$.

```{r}
fit_result <- model_fit(
  response ~ location + f(
    smoothing_var = floor.size,
    model = "IWP", k = 10,
    order = polyOrder1, initial_location = NULL,
  )
  + score + f(
      smoothing_var = year,
      model = "IWP",
      order = polyOrder2, k = 30, # should add a checker for k >= 3
      sd.prior = list(prior = "exp", para = list(u = 1, alpha = 0.5)),
      boundary.prior = list(prec = 0.01), initial_location = NULL
    ),
  data = data, method = "aghq", family = "Binomial",
  control.family = list(sd_prior = list(prior = "exp", para = list(u = 1, alpha = 0.5))),
  control.fixed = list(intercept = list(prec = 0.01), location = list(prec = 0.01), score = list(prec = 0.01))
)
```

```{r}
names(fit_result)
IWP1 <- fit_result$instances[[1]]
IWP2 <- fit_result$instances[[2]]
mod <- fit_result$mod

slotNames(IWP1)
names(mod)
```
### Let's take out some samples:
```{r}
samps <- aghq::sample_marginal(mod, M = 3000)

global_samps1 <- samps$samps[fit_result$boundary_samp_indexes$floor.size, , drop = F]
coefsamps1 <- samps$samps[fit_result$random_samp_indexes$floor.size, ]

global_samps2 <- samps$samps[fit_result$boundary_samp_indexes$year, , drop = F]
coefsamps2 <- samps$samps[fit_result$random_samp_indexes$year, ]

fixed_samps <- samps$samps[fit_result$fixed_samp_indexes$Intercept, , drop = F]

f1 <- compute_post_fun(
  samps = coefsamps1, global_samps = global_samps1,
  knots = IWP1@knots,
  refined_x = seq(min(IWP1@observed_x), max(IWP1@observed_x), length.out = 100),
  p = polyOrder1, degree = 0
)

f2 <- compute_post_fun(
  samps = coefsamps2, global_samps = global_samps2,
  knots = IWP2@knots,
  refined_x = seq(min(IWP2@observed_x), max(IWP2@observed_x), length.out = 100),
  p = polyOrder2, degree = 0
)

f1pos <- extract_mean_interval_given_samps(f1)
f2pos <- extract_mean_interval_given_samps(f2)


matplot((f1pos$x + IWP1@initial_location), f1pos[, c("mean", "plower", "pupper")], lty = c(1, 2, 2), lwd = c(2, 1, 1), col = "black", type = 'l', ylab = "effect on rent", xlab = "floor.size")

matplot((f2pos$x + IWP2@initial_location), f2pos[, c("mean", "plower", "pupper")], lty = c(1, 2, 2), lwd = c(2, 1, 1), col = "black", type = 'l', ylab = "effect on rent", xlab = "year")
lines(I(func(f2pos$x + IWP2@initial_location) - func(f2pos$x + IWP2@initial_location)[1]) ~ I(f2pos$x + IWP2@initial_location), col = "red")

fixed_samps %>% apply(1, mean)
fixed_samps %>% apply(1, sd)
```
